---
title: "From lab spreadsheets to maps with geochemR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{From lab spreadsheets to maps with geochemR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 8, fig.height = 5, dpi = 90,
                      fig.alt = "geochemR plot of the synthetic example dataset")
```

```{r setup}
library(geochemR)
```

## 1. Reading what the lab sent

Laboratories deliver *wide* tables — one row per sample, one column per
mineral, oxide or pyrolysis parameter — with their own spellings and unit
conventions. The readers accept that as-is:

```{r read}
xrd <- data.frame(Sample = c("A1", "A2"), Quartz = c(40, 12), "K-Feldspar" = c(5, 1),
                  "Illite/Mica" = c(25, 3), "I/S" = c(6, 1), Calcite = c(20, 80), Pyrite = c(2, 1),
                  "Total Clay (wt%)" = c(31, 4), check.names = FALSE)
read_xrd(xrd, lab = "Lab A")

xrf <- data.frame(Sample = c("A1", "A2"), SiO2 = c(62.1, 8.3), CaO = c(9.9, 48.2),
                  Zr_ppm = c("180", "<5"), "Ba (ppm)" = c(410, 95), check.names = FALSE)
read_xrf(xrf, lab = "Lab B")
```

`"<5"` became `value = 5, lod = 5, qualifier = "<"`. Units were taken from
the column suffixes. Column names that are not analytes are reported, not
silently dropped.

A sample table gives every `sample_id` a hole, coordinates and a depth
interval; `gc_data()` puts the pieces together and validates them.

```{r model}
samples <- data.frame(sample_id = c("A1", "A2"), hole_id = "W1", x = 500100, y = 4200100,
                      depth_top = c(50, 60), depth_base = c(60, 70))
ds <- gc_data(samples, dplyr::bind_rows(read_xrd(xrd), read_xrf(xrf)), crs = 26914)
ds
gc_wide(ds, "XRD")
```

## 2. The example dataset

`gc_example` is synthetic: twelve holes on a grid, fifteen 10-ft intervals
each, through a sandstone, an organic-rich calcareous mudstone whose TOC
rises eastward, and a limestone. Every sample has XRD and XRF; core and most
cuttings have pyrolysis.

```{r example}
ds <- gc_example
ds
gc_analytes(ds)
```

## 3. Processing

Substitute censored values, add the derived indices, and — for anything
statistical on compositions — take log-ratios.

```{r process}
ds <- ds |>
  gc_substitute_lod("half") |>
  gc_indices()
gc_analytes(ds)$SRA

w <- gc_wide(ds, "XRD", c("quartz", "carbonate", "clay"))
head(gc_clr(w[, c("quartz", "carbonate", "clay")]))
```

## 4. Depth

```{r depth, fig.height = 5}
plot_depth_profile(ds, "SRA", c("TOC", "S2", "Tmax", "HI"), holes = c("H01", "H04", "H09", "H12"))
plot_depth_heatmap(ds, "SRA", "TOC")
plot_section(ds, "SRA", "TOC", holes = c("H01", "H02", "H03", "H04"))
plot_mineralogy(ds, holes = c("H01", "H12"))
```

## 5. Plan

`plot_map()` summarises each hole over a depth window and draws an
inverse-distance surface under the points.

```{r map, fig.width = 6, fig.height = 5.5}
plot_map(ds, "SRA", "TOC", depth = c(50, 110))
plot_map(ds, "XRF", "Mo", depth = c(50, 110), fun = max)
```

## 6. Cross-plots

```{r cross, fig.width = 6, fig.height = 5}
plot_kerogen(ds, "hi_oi")
plot_kerogen(ds, "hi_tmax", colour = "depth_mid")
plot_ternary(ds, "XRD", c("quartz", "carbonate", "clay"), colour = "depth_mid")
```

## 7. A whole laboratory workbook at once

Commercial cuttings-analysis deliverables put every method on one sheet
under a group / analyte / unit header. `read_workbook()` reads that layout
directly; the package ships a synthetic one in exactly that form.

```{r workbook}
f <- system.file("extdata", "cuttings_workbook.xlsx", package = "geochemR")
cu <- read_workbook(f, hole_id = "EX-1", lab = "Example Cuttings Lab") |> gc_indices()
cu
gc_analytes(cu)$PAM
```

Bottom-only depths become intervals, formation and zone come along, the clay
speciation (reported as % of total clay) is converted to bulk wt%, and the
lab's own calculated columns are dropped so that `gc_indices()` is the one
source of derived values.

```{r cuttings-plots, fig.height = 6}
plot_depth_profile(cu, "SRA", c("TOC", "S1", "S2", "Tmax", "HI"))
plot_pam(cu)
plot_kerogen(cu, "hi_tmax", colour = "formation")
gc_hole_summary(cu, "SRA", by = c("hole_id", "formation"), analytes = c("TOC", "HI", "Tmax", "Ro_eq"))
```

## 8. The look

Every plot uses `theme_gc()` — parchment page, brown ink, serif type,
tan-to-espresso palette — and the `scale_*_gc()` scales. They are ordinary
ggplot2 objects, so add layers or override anything:

```{r look}
plot_map(ds, "SRA", "TOC", depth = c(50, 110)) +
  ggplot2::labs(title = "Where the organic-rich mudstone is richest",
                caption = "Synthetic example; mean TOC per hole, inverse-distance surface")
```

## 9. Summaries for reports

```{r summaries}
gc_hole_summary(ds, "SRA", depth = c(50, 110), analytes = c("TOC", "HI", "Tmax", "Ro_eq"))
head(gc_interval_stats(ds, "XRD", breaks = 50, analytes = c("quartz", "carbonate", "clay", "BI_min")))
```
